---
title: "Working with Maps and GIS data"
#author: "Liming Wang, Joe Broach"
#date: "8/21/2017"
#updated: "2019-04-09"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)

```

# Notable packages

## The tidycensus packages
- Website: https://walkerke.github.io/tidycensus/
- Install with `install.packages("tidycensus")`
- Vignettes: https://walkerke.github.io/tidycensus/articles/basic-usage.html
- Reference: https://walkerke.github.io/tidycensus/reference/index.html

## The simple feature packages
- Website: https://github.com/r-spatial/sf
- Install with `install.packages("sf")`
    - On Mac, you may need to install `udunits` first, for example, with [homebrew](https://stackoverflow.com/a/42387825/688693)
    - On Linux, install `libudunits2-dev` with your software management system first
- Vignettes: https://r-spatial.github.io/sf/articles/sf1.html
- Reference: https://r-spatial.github.io/sf/reference/index.html


# Code examples

## Data files used in the examples 

1. [Portland Metro 1994 TAZ shape file](data/taz1260.shp.zip): Download & unzip to `data` subdirectory of your RStudio project folder
2. [Portland Metro 1994 Survey geocode.raw.zip](data/portland94_geocode.raw.zip)

## Map ACS 2013-17 Median Household Income by Census Tract

```{r}
## Install the tidycensus package if you haven't yet
install.packages("tidycensus")
install.packages("dplyr")
install.packages("lwgeom")

library(dplyr)
library(tidycensus)
library(ggplot2)
library(tidyr)
library(lwgeom)
library(units)

## setup cenus api key
## signup your census api key at http://api.census.gov/data/key_signup.html
census_api_key("5d1693be6ce12231248f292759f241c1ead3ed53") # note: needed to re-build website
v17 <- load_variables(2017, "acs5", cache = TRUE)
View(v17)
pdx_tracts_raw <- get_acs(geography = "tract", 
                        year = 2016, # 2013-2017
                        variables = c("B19013_001",
                                      "B03002_003",
                                      "B01003_001"),
                        state = "OR", 
                        county = c("Multnomah County"),
                        geometry = TRUE) %>% # load geometry/gis info
  select(GEOID, variable, estimate, geometry)

pdx_tracts <- pdx_tracts_raw %>%
  spread(key=variable, value=estimate) %>%
  rename(medinc = B19013_001, white = B03002_003, pop = B01003_001) %>%
  mutate(nwnh_pct = (pop - white) / pop, area = st_area(geometry))
head(pdx_tracts)
pdx_tracts_p <- sf::st_transform(pdx_tracts, crs = 2838) %>%
  mutate(area = st_area(geometry))
head(pdx_tracts_p)
plot(pdx_tracts_p)
plot(pdx_tracts)

pdx_tracts_p %>%
  arrange(desc(area))

pdx_tracts_p <- pdx_tracts_p %>%
  filter(area < units::as_units(1e8, "m^2"))
?select
(myplot <- ggplot(pdx_tracts_p) + 
  geom_sf(aes(fill = medinc)) +
  theme_minimal())

biketown <- read.csv("biketown_all.csv",

                                          stringsAsFactors = F)
head(biketown)


hubs_start_sf <- biketown %>%
  group_by(StartHub) %>%
  summarise(lat = mean(StartLatitude), lng = mean(StartLongitude),
            starts = n()) %>%
  filter(!is.na(lat)) %>%
  st_as_sf(coords = c("lng", "lat"), 
           crs = 4326, agr = "constant")

hubs_start_sf <- st_transform(hubs_start_sf, 2838)
hubs_start_sf_j <- hubs_start_sf %>%
  sf::st_join(pdx_tracts_p)
head(hubs_start_sf_j)
#ggsave("output/mymap.pdf", myplot)

```

## Interactive Maps of ACS 2012-16 Median Household Income by Census Tract

```{r, results="asis"}
## Install the mapview package if you haven't yet
#install.packages("mapview")
library(sf)
library(mapview)
library(dplyr)

mapview(portland_tract_medhhinc %>% select(estimate), 
        col.regions = sf.colors(10), alpha = 0.1)
```

## Example of spatial analysis: spatial join

```{r}
library(sf)
library(readr)
# read 1994 Metro TAZ shape file
taz_sf <- st_read("data/taz1260.shp", crs=2913)

# read geocode.raw file that contains X and Y coordinates
portland94_df <- read_csv("data/portland94_geocode.raw.zip", col_names=c("uid", "X", "Y", "case_id", 
                                                                       "freq", "rtz", "sid", 
                                                                       "totemp94", "retemp94"))
portland94_df <- portland94_df %>% 
  filter(X!=0, Y!=0) %>% 
  sample_n(500)

# create a point geometry with x and y coordinates in the data frame
portland94_sf <- st_as_sf(portland94_df, coords = c("X", "Y"), crs = 2913)

# spatial join to get TAZ for observations in portland94_sf
portland94_sf <- st_join(portland94_sf, taz_sf)
head(portland94_sf)

ggplot() +
  geom_sf(data = taz_sf, aes(alpha=0.9)) +
  geom_sf(data = portland94_sf)
```

# Resources:
- [R spatial CRAN Task View](https://cran.r-project.org/web/views/Spatial.html)
